---
title: "Eury_loxo_SEI_model"
author: "Alyssa-Lois Gehman"
date: "11/20/2017"
output:
  html_document:
    fig_height: 8
    fig_width: 9
  pdf_document:
    fig_height: 8
    fig_width: 9
---

#Host and parasite thermal ecology jointly determine the effect of climate warming on epidemic dynamics 
Authors:  Alyssa-Lois M. Gehman, Richard J. Hall and James E. Byers
Correspondence to: 	Alyssa-Lois M. Gehman

#This code produces the seasonal host-parasite dynamics used in manuscript, based on weekly temperature data

*Requirements*
  a. download and install package deSolve
  b. download all files to a single folder and open file 'Gehman_codeshare.Rproj'

*1 - Functional fits to the data*
These define the best fit model parameters for the temperature dependent rates in the SEI mode, as parameterized from the temperature data, and recreate Figure 1C-E.


```{r briere fit, echo=FALSE}

#make sure this file and all data are in same directory 


# 1a. function for the Briere (Eq. 2, fun.b) and modified Briere (Eq. 3, fun.mb) 
#     used to plot the best fit models
      #Input variables c, T0 (=Tmin), Tm (=Tmax), and m are estimated from the nls best fit (1b)
      # h.temp is a number series that encompases the range of temperature values 
      # you want to plot over 
      fun.mb=function(c,T0,Tm,h.temp,m){c*(-h.temp+T0+Tm)*(Tm-h.temp)*(h.temp-T0)^(1/m)}
      fun.b=function(c,T0,Tm,h.temp,m){c*h.temp*(h.temp-T0)*(Tm-h.temp)^(1/m)}

#1b. The data for parasite reproduction over temperature fit to a modified Briere (Fig. 1B).
#     The output of the nonlinear least squares fit gives estimates of c, T0, Tm.  
      # read in data for parasite reproduction across temperature
      beta.data<-read.csv("parasite_rep.csv")  
      #add the path for the directory you are working in to access this data

      # The best fit nonlinear least squares model with 1/m=0.65
      rmb<- nls(lxmx ~ c*(-temp+T0+Tm)*(Tm-temp)*(temp-T0)^(0.65), data=beta.data, 
                start=list(c=12.083, T0=9.875, Tm=30.753))
  
      # temperature simulations - set to the range of values that you want to plot 
      # the modified Briere fit over
      tempsim=seq(1, 35, by=.01)
      # Figure 1B - plotting the mean parasite reproductive output by temperature 
      plot(beta.data$temp, beta.data$lxmx, 
           pch=16, ylim=c(0, 25000), xlim=c(0,35),
           xlab="Temperature", ylab="Parasite Reproduction (lxmx)", col="deeppink", axes=F)
      #plotting error bars of the standard error of parasite reproductive output
      segments(beta.data$temp, y0=beta.data$lxmx.neg, y1=beta.data$lxmx.pos,
               col="deeppink", lwd=2)
      #plotting the modified Briere fit with c=12.083, T0=9.875, Tm=30.753,(from rmb) 
      # and across temperature range (tempsim)
      lines(tempsim, fun.mb(12.083, 9.875,30.753,tempsim,1/0.65),
            lwd=3, col="deeppink") #rmb - model the estimates came from
      #plotting the x-axis
      axis(1, c(0,5,10,15,20,25,30,35), c(0,5,10,15,20,25,30,35)) 
      #plotting the y-axis
      axis(2, c(0,5000, 10000, 15000, 20000))

      #The output with the estimates for c, T0 and Tm
      rmb


#1.c The data for susceptible expected lifespan over temperature fit to a Briere (Fig. 1C). The output of the nonlinear least squares fit gives estimates of c, T0, Tm.   

      #read in data for susceptible expected lifespan over temperature
      S.data=read.csv("Susceptible_lifespan.csv") #add the path for the directory you are working in to access this data


      # The best fit nonlinear least squares model with 1/m=2
      hb1 <- nls(S.surv~c*S.temp*(S.temp-T0)*(Tm-S.temp)^(2), data=S.data, start=list(c=0.0008336, T0=4.05, Tm=33.85))

      # temperature simulations - set to the range of values that you want to plot the modified Briere fit over
      sus.temp=seq(2.29,35,1)

      #Figure 1C - plotting the mean susceptible expected lifespan
      plot(S.data$S.temp, S.data$S.surv,
           pch=16, ylim=c(0,60), xlim=c(0,35),axes=F,
           xlab="Temperature", ylab="Expected Lifespan (Weeks)", col="green")
      #plotting error bars of the standard error of susceptible host expected lifespan
      segments(S.data$S.temp, y0=S.data$S.surv-S.data$S.surv.se, y1=S.data$S.surv+S.data$S.surv.se, col="green")
      #plotting the Briere fit with c=0.0006923, T0=3.136, Tm=34.8, (from hb1) and across temperature range (sus.temp)
      lines(sus.temp, fun.b(0.0006923,3.136,34.8,sus.temp,1/2), lty=1, lwd=3, col="green")
      #plotting the x-axis
      axis(1, c(0,5,10,15,20,25,30,35), c(0,5,10,15,20,25,30,35))
      #plotting the y-axis
      axis(2, c(0,20,40,60))
      
      #The output with the estimates for c, T0 and Tm
      hb1
      
# 1d. The data for exposed expected lifespan over temperature fit to a modified Briere (Fig. 1D). The output of the nonlinear least squares fit gives estimates of c, T0, Tm. 
      ## read in the data for exposed expected lifespan over temperature
      E.data=read.csv("Exposed_lifespan.csv") #add the path for the directory you are working in to access this data

      # The best fit nonlinear least squares model with 1/m=0.4
      vmb1 <- nls(E.surv ~ c*(-E.temp+T0+Tm)*(Tm-E.temp)*(E.temp-T0)^(0.4), data=E.data, start=list(c=.028, T0=4.9, Tm=36))

      # temperature simulations - set to the range of values that you want to plot the modified Briere fit over
      e.temp=seq(4,35,0.01)

      #Figure 1D - plotting the mean exposed expected lifespan
      plot(E.data$E.temp, E.data$E.surv,
           pch=16, ylim=c(0,60), xlim=c(0,35), axes=F,
           xlab="Temperature", ylab="Expected Lifespan (Weeks)", col="purple")
      #plotting error bars of the standard error of exposed host expected lifespan
      segments(E.data$E.temp, y0=E.data$E.surv-E.data$E.surv.se, y1=E.data$E.surv+E.data$E.surv.se, col="purple")
      #plotting the modified Briere fit with c=0.03393, T0=4.95039, Tm=34.43724 (from vmb1) and across temperature (e.temp)
      lines(e.temp,fun.mb(0.03393,4.95039,34.43724,e.temp,1/0.4), lwd=3, col="purple")
      #plotting the x-axis
      axis(1, c(0,5,10,15,20,25,30,35), c(0,5,10,15,20,25,30,35))
      #plotting teh y-axis
      axis(2, c(0,20,40,60))
      
     #The output with the estimates for c, T0 and Tm
      vmb1

#1e. The data for infected expected lifespan over temperature fit to a modified Briere (Fig. 1E). The output of the nonlinear least squares fit gives estimates of c, T0, Tm. 
      ## read in data for infected expected lifespan over temperature
      I.data=read.csv("Infected_lifespan.csv") #add the path for the directory you are working in to access this data

      # The best fit nonlinear least squares model with 1/m=0.3
      imb <- nls(I.surv ~ c*(-I.temp+T0+Tm)*(Tm-I.temp)*(I.temp-T0)^(0.3), data=I.data, start=list(c=0.04347, T0=4.99901, Tm=31.85831))
      
      #temp for plotting Briere
      i.temp=seq(0,31.29,0.05)

      #Figure 1E - plotting the mean expected infected lifespan over temperature
      plot(I.data$I.temp, I.data$I.surv,
           pch=16, ylim=c(0,60), xlim=c(0,35),axes=F,
           xlab="Temperature", ylab="Expected Lifespan (Weeks)", col="orange")
      #plotting error bars of the standard error of infected host expected lifespan
      segments(I.data$I.temp, y0=I.data$I.surv-I.data$I.surv.se, y1=I.data$I.surv+I.data$I.surv.se, col="orange")
      #plotting the modified Briere fit with c=.03308, T0=4.94846, Tm=31.29617 (from imb) and across temperature (i.temp)
      lines(i.temp, fun.mb(.03308,4.94846,31.29617,i.temp,1/0.5), lwd=3, col="orange")
      #plotting x-axis
      axis(1, c(0,5,10,15,20,25,30,35), c(0,5,10,15,20,25,30,35))
      #plotting y-axis
      axis(2, c(0,20,40,60))
      
      #The output with the estimates for c, T0 and Tm
      imb


```


*2 - The model equations*

Must be run for sections 3 and 4 to work. 

```{r rhizo sei model, echo=FALSE}

#library required to run the model
require(deSolve)

#remove all lists from previous runs
rm(list=ls(all=TRUE)) 

#THE MODEL (Eq 12,13 and 14):
rhizo_sei <- function(t,y,params) {
  S = y[1]
  E = y[2]
  I = y[3]
  
  with(as.list(c(params)), {
    dS <- delta - (mu_s * S) - (beta * S * I)
    
    dE <- (beta * S * I) - (tau * E) - (mu_e * E)
    
    dI <- (tau * E) - (mu_i * I)
    
    dy = c(dS,dE,dI)
    list(dy)
  })
}


### functions from Briere Fits
fun.mb=function(c,T0,Tm,h.temp,m){c*(-h.temp+T0+Tm)*(Tm-h.temp)*(h.temp-T0)^(1/m)}
fun.b=function(c,T0,Tm,h.temp,m){c*h.temp*(h.temp-T0)*(Tm-h.temp)^(1/m)}


```

*3 - Evaluating climate change effects with the model*

Below we use thermal performance curves (Section 1) to parameterize how host mortality and transmission rates depend on weekly mean temperatures; the other parameters were estimated from the literature, or chosen to give good quantitative agreement with field data on host abundance and infection prevalence by season and subject to extensive sensitivity analyses (methods C.ii and iii). 

The temperatures used are mean weekly temperatures from a mooring that is part of the Georgia Coastal Ecosystems Long Term Ecological Research program (GCE-LTER) that measured temperature at a depth of ~1m (temp0), the approximate depth of water found over an oyster reef in Georgia, and thus the thermal environment experienced by E. depressus and L. panopaei.

The model runs with ambient temperature (temp0), then ambient +1 and +2ºC (temp1 and temp2), then the mean temperature change from 1970-2008 (temp1.6) and seasonally varying change (temp.season) from 1970-2008 (tempseas). 

To run this code, first run the code in section 2.


```{r rhizo sei model - varying temp, echo=FALSE, warning=FALSE}

rm(total_prev) ; rm(pop); rm(pop_e_temp); rm(pop_i_temp); rm(pop_s_temp) # remove variables from previous runs


### Temperature data - must be in the form of mean weekly temperatures - 1 mean temperature for each of 52 weeks.

# GCE data - load temperature data
gce10.data=read.csv("gce10.wk.mean.csv") #add the path for the directory you are working in to access this data
#create a variable with the mean weekly temperatures in it
gce10.wk.mean=gce10.data$V1

# Set seasonally varying climate change
temp.season=c(rep(1.5,11),rep(0.67,13),rep(0.89,13),rep(0.61,13),rep(1.5,2)) # based on 1970-2008, Carl, Mellilo and Peterson 2009

# Set up climate change scenarios
temp0=data.frame(wk=gce10.wk.mean[1:52]) #ambient temperature
temp1=data.frame(wk=gce10.wk.mean[1:52]+1) #ambient +1ºC temperature
temp2=data.frame(wk=gce10.wk.mean[1:52]+2)  #ambient +2ºC temperature
temp0.89=data.frame(wk=gce10.wk.mean[1:52]+0.89) #ambient +0.89ºC temperature
tempseas=data.frame(wk=gce10.wk.mean[1:52]+temp.season) #ambient seasonally varying temperature change

## Combine the climate change scenarios into a data frame
tempramp=data.frame(temp0,temp1,temp2,temp0.89,tempseas)

### Running model with varying temperature based on GCE annual data 

#THE PARAMETERS:
times = seq(0,1, by=0.1) 
tau = 0.04       # tau set to 25 weeks (1/25=0.04)

ystart = c(S=5000,E=10,I=2000) # The starting values for the Susceptible (S), Exposed (E) and Infected (I)

nyears=10        # how many years we want to run the model for
nweeks=52*nyears     # translating year to weeks

#creating empty vectors to fill with model output
total_prev=matrix(rep(0, length(tempramp)*length(1:nweeks)), nrow=length(1:nweeks)) #prevalence by week 
pop_i_temp=matrix(rep(0, length(tempramp)*length(1:nweeks)), nrow=length(1:nweeks)) #infected population by week
pop_e_temp=matrix(rep(0, length(tempramp)*length(1:nweeks)), nrow=length(1:nweeks)) #exposed population by week
pop_s_temp=matrix(rep(0, length(tempramp)*length(1:nweeks)), nrow=length(1:nweeks)) #suscpeptible population by week
mu_i_temp=matrix(rep(0, length(tempramp)*length(1:nweeks)), nrow=length(1:nweeks)) #infected mortality rate by week (1/expected lifespan)
mu_e_temp=matrix(rep(0, length(tempramp)*length(1:nweeks)), nrow=length(1:nweeks)) #exposed moratlity rate by week (1/expected lifespan)
mu_s_temp=matrix(rep(0, length(tempramp)*length(1:nweeks)), nrow=length(1:nweeks)) #susceptible mortality rate by week (1/expected lifespan)
beta_time=matrix(rep(0, length(tempramp)*length(1:nweeks)), nrow=length(1:nweeks)) #parasite reproduction by week

#set up forloop to cycle through the temperature scenarios
for(j in 1:length(tempramp)){
   tempset=tempramp[j]  #input the temperatures to run
   ystart = c(S=5000,E=10,I=2000)
   #set up a forloop to cycle through the weeks in the model
    for(i in 1:nweeks){
       # weekly variance set up 
        wk=i %% 52                      # Get the week of the year  
        wk.fix=if (wk==0) +1 else wk    # Fix the 0 week and make it a 1
        temp=tempset[wk.fix,]     # call the temp for each week (from four years of weekly averaged gce10 temps)
        # functional fit - plus a correction with mortality set at a relative maximum when the model goes above the thermal limit of susceptible mortality (34.8ºC)
        mu_s=if (temp>34.8) 6 else 1/fun.b(0.0006923,3.136,34.8,temp, 1/2) 
        # function fit for exposed mortality with a correction set at relative maximum when the model goes above the thermal limit of exposed mortality (34.43)
        mu_e=if (temp>34.43) 6 else 1/fun.mb(0.03393,4.95039,34.43724,temp,1/0.4) 
        # functional fit - plus a correction with mortality set at a relative maximum when the model goes above the thermal limit of infected mortality (31.104ºC)
        mu_i=if (temp>31.104) 6 else 1/fun.mb(0.04414, 4.99944, 32.10496,temp, 1/0.3) 
        # Host recruitment is turned on between 32 and 38 weeks
        delta=if (wk.fix>31 & wk.fix<37) 2400 else 0 
        # functional fit - plus a correction with mortality set at a relative maximum when the model goes above the thermal limit of infected mortality (30.75ºC)
        beta=if (temp>30.75) 0 else fun.mb(0.0000004, 9.875,30.753, temp, 1/0.65)   
        # a string of the parameters set by temperature
        params=c(beta, mu_s, mu_e, mu_i, delta, tau)
            # solve equation
            outd = lsoda(ystart,times,rhizo_sei,params) # times is between 0 and 1
            # output population for each week
            Sfinal = tail(outd,n=1)[2]
            Efinal = tail(outd,n=1)[3]
            Ifinal = tail(outd,n=1)[4]
            # output mortality rates for each week
            mu_sfinal= tail(mu_s)
            mu_efinal= tail(mu_e)
            mu_ifinal= tail(mu_i)
            # oubtput beta for each week
            beta_ifinal=tail(beta)
           # write prevalence to i,j entry of matrix
            total_prev[i,j]=Ifinal/(Sfinal+Efinal+Ifinal) 
            #write population to i,j entry of matrix
            pop_s_temp[i,j]=Sfinal
            pop_e_temp[i,j]=Efinal
            pop_i_temp[i,j]=Ifinal
            ## write mu to i,j entry of the matrix
            mu_s_temp[i,j]=mu_sfinal
            mu_e_temp[i,j]=mu_efinal
            mu_i_temp[i,j]=mu_ifinal
            ## write beta to i,j entry of the matrix
            beta_time[i,j]=beta_ifinal
            # reset ystart with the output from that week
            ystart=c(Sfinal, Efinal, Ifinal)
      } # end i loop
} # end j loop


```


*4 Plotting climate change effects*

This code using the model output from section 3 and creates Figure 3.  For this code to work first section 2 and 3 must have been run.

```{r - climate change plots, echo=FALSE, }

## graphic settings 
week=c(1:52) #weeks in a year
m.tic=c(2,6,10,15,20,24,28,32,36,41,45,50) #tick locations for month labeling
m.lab=c("Jan", "", "Mar", "", "May", "", "July", "", "Sept", "", "Nov", "") #month labeling
col_pap=c("#fdcc8a", "#fc8d59", "#d7301f","#df65b0",  "#980043") #colors for the graphics


par(mfrow=c(3,2)) #set up 3 rows and 2 columns of graphs
############################################################
#########  first row - water temp    #####################
############################################################
### [1,1] amb +CC
  par(mar=c(2.1,4.1,2.1,2.1)) #Settings for the margins that arrange multiple graphs within one page 
  plot("","", xlim=c(0,52), ylim=c(10,35), 
        axes=F, xlab="", ylab="Temperature (ºC)", main="Annual Mean Temperature Increase", cex.lab=1.4)
          for(i in 1:3){
              lines(week, tempramp[1:52,i], col=col_pap[i],lwd=3) #loop with temperature scenario 1-3
              }
        axis(2, seq(10,35, 5), cex.axis=1.4) #y axis with prevalence
        axis(1,m.tic, labels=m.lab, cex.axis=1.4) # x-axis with months
        text(2,34,"A", cex=1.4) # The letter label
        text(1,31, "Ambient +2ºC", col="#d7301f", pos=4, cex=1.6) # label for ambient +2ºC tempurature
        text(1,29,"Ambient +1ºC", col="#fc8d59", pos=4, cex=1.6) # label for ambient +1ºC tempurature
        text(1,27, "Ambient", pos=4, cex=1.6, col="#fdcc8a") # label for ambient temperature

## [1,2] amb +seas CC
  par(mar=c(2.1,4.1,2.1,2.1))  #Settings for the margins that arrange multiple graphs within one page 
  plot("","", xlim=c(0,52), ylim=c(10,35), 
       axes=F, xlab="", ylab="", main="Seasonal Mean Temperature Increase", cex.lab=1.4)
       lines(week, tempramp[1:52, 1], lwd=3, col="#fdcc8a")     # ambient temperature scenario
          for(i in 4:5){
              lines(week, tempramp[1:52,i], col=col_pap[i],lwd=3) #loop with temperature scenario 4-5
             }
       axis(2, seq(10,35, 5), cex.axis=1.4) #y axis with prevalence
       axis(1,m.tic, labels=m.lab, cex.axis=1.4) # x-axis with months
       text(2,34,"B", cex=1.4) # The letter label
       text(1,31,"Ambient +SeasºC", col="#980043", pos=4, cex=1.6) # label for seasonally varying temperature
       text(1,27, "Ambient", pos=4, cex=1.6, col="#fdcc8a") # label for ambient temperature
       text(1,29, "Ambient +1.6ºC", col="#df65b0", pos=4, cex=1.6) # label for ambient +1.6ºC temperature

############################################################
#########  second row -  prevalence   #####################
############################################################

## [2,1] prev by ambient +CC
  par(mar=c(4.1,4.1,0.1,2.1)) #Settings for the margins that arrange multiple graphs within one page
  plot("","", xlim=c(0,52), ylim=c(0,60), 
      axes=F, xlab="", ylab="Infection Prevalence (%)", cex.lab=1.4)
        for(i in 1:3){
            lines(week, total_prev[469:520,i]*100, pch=16, col=col_pap[i], lwd=3) #loop with temperature scenario 1-3
            }
      axis(2, seq(0,60, 25), cex.axis=1.4) #y axis with prevalence
      axis(1,m.tic, labels=m.lab, cex.axis=1.4) # x-axis with months
      text(2, 50, "C", cex=1.4) # The letter label
## [2,2] prev by ambient +seas
  plot("","", xlim=c(0,52), ylim=c(0,60), axes=F, xlab="", ylab="")
      lines(week, total_prev[469:520,1]*100, lwd=3, col="#fdcc8a") # ambient temperature scenario
      for(i in 4:5){
          lines(week, total_prev[469:520,i]*100, pch=16, col=col_pap[i], lwd=3) #loop with temperature scenario 4-5
          }
      axis(2, seq(0,60, 25), cex.axis=1.4) #y axis with prevalence
      axis(1,m.tic, labels=m.lab, cex.axis=1.4) # x-axis with months
      text(2, 50, "D", cex=1.4) # The letter label
############################################################
#########  third row - total population      #####################
############################################################

## [3,2] SEI pop ambient +CC
  par(mar=c(4.1,4.1,0.1,2.1))
  plot("","", xlim=c(0,52), ylim=c(0,20000), axes=F, xlab="", ylab="Total Host Abundance", cex.lab=1.4)
    for(i in 1:3){
        totpop=pop_s_temp[469:520,i]+pop_e_temp[469:520,i]+pop_i_temp[469:520,i] #sum the population output
        lines(week, totpop, col=col_pap[i], lwd=3)  #loop with temperature scenario 1-3
        }
      axis(2, seq(0,20000, 20000/2), cex.axis=1.4) #y axis with total host abundance    
      axis(1,m.tic, labels=m.lab, cex.axis=1.4) # x-axis with months
      text(2, 19000, "E", cex=1.4) # The letter label
      mtext("Date",side=1,col="black",line=3) #label the x-axis


## [3,3] SEI pop ambient +CC
  plot("","", xlim=c(0,52), ylim=c(0,20000), axes=F, xlab="", ylab="")
    lines(week, pop_s_temp[469:520,1]+pop_e_temp[469:520,1]+pop_i_temp[469:520,1], lwd=3, col="#fdcc8a") # ambient temperature scenario
    for(i in 4:5){
        totpop=pop_s_temp[469:520,i]+pop_e_temp[469:520,i]+pop_i_temp[469:520,i] #sum the pouplation output
        lines(week, totpop, pch=16, col=col_pap[i], lwd=3) #loop with temperature scenario 4-5
        }
    axis(2, seq(0,20000, 20000/2), cex.axis=1.4) #y axis with total host abundance
    axis(1,m.tic, labels=m.lab, cex.axis=1.4) # x-axis with months
      text(2, 19000, "F", cex=1.4) # The letter label
      mtext("Date",side=1,col="black",line=3) #label the x-axis


```



